Dark Control:\\ A Unified Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-05-06 paper/paper.tex
more more stuff
major pass: link between fep and rl

\begin{equation}
  \pi^{\text{greedy}}(s) = \argmax_{a \in \mathcal A}\tilde{Q}(s, a|\theta),
  \label{eq:policy}
\end{equation}
This defines a mapping from states directly to actions.
% $a \mapsto \tilde{Q}(a,s|\theta)$ of the $Q$-value function ``frozen'' at state $s$.
For instance, a simple linear model with a kernel $\phi$ would be of the
form $\tilde{Q}(s, a|\theta) = \phi(s,a)^T\theta$, where
$\phi(s,a)$ would represent a high-level representation of the state-action pairs
$(s,a)$, as was done proposed \citep{songNIPS2016}, or a eural network as recently proposed by in the
breakthrough works
\citep{mnih2015,silver2016mastering} for playing complex games (atari, Go, etc.) at superhuman levels.
%  and Parr et al. 2003 (JMLR)].
In the DMN, the dmPFC would implement such a hard-max lookup
over the action space.
The model
  parameters $\theta$ would correspond to synaptic weights and connection strengths between
  brain regions. It is a time-varying neuronal program which dictates how to move from world states $s$ to actions $a$ via the hard-max policy \eqref{eq:policy}.
  The approximating $Q$-value function $\tilde{Q}(s, a|\theta)$ would tell the DMN the (expected) usefulness of taking an action $a$ in state $s$.
  The DMN, and in particular its dmPFC node, could then contribute to the choice, at a given state $s$, of an action $a$ which maximizes these approximate
  $Q$-values. Such a mapping from states to actions is conventionally called \textit{policy matrix}\citep{mnih2015,silver2016mastering}.
  Learning consists in starting from a given table and
  updating it during action choices,
  which take the agent to different table entries.

@@

\subsection{The free-energy principle and active inference}
\textbf{XXX Passage: Explain FEP and active inference}
In Friston's free-energy principle (FEP) \citep{friston2010free,fristonAIorRL}, the brain is portrayed
as biomechanical inference engine
dedicated to minimizing the long-term average of surprise.
Precursors of this theory can be traced back to \citep{dayan1995helmholtz} in which they introduced the so-called \textit{Helmholtz machine}, a hierarchical factorial \textit{directional deep belief-net}. According to the FEP account,
the goal of the brain is then to optimize over a generative model $G$ of sensations: to iteratively
modify its internal representation $p_G(\z|\x)$ about objects in the world, their interactions and dynamics,
so as to minimize surprise when these representations are confronted with sensory input $\x$ during perception cycles (i.e., the \textit{generative} model).
The FEP also postulates a dual model called the \textit{recognition} model,
which works in tandem with the generative model $p_R(\z|\x)$, to accomplish approximate inference. The recognition model dreams imaginary worlds $\z$,
while the latter tries to generate sensations $\x$ which match these dreams!

@@

% \begin{figure}[!tb]
%   \centering
%   \includegraphics[width=.8\linewidth]{vae.pdf}
%   \caption{Variational autoencoders...}
%   \label{fig:vae}
% \end{figure}
\textbf{XXX Passage: Explain the relation of FEP and active inference to our MDPs}
The main algorithm for minimizing free-energy $F_G^R(\x)$ is the \textit{wake-sleep algorithm}
\citep{dayan1995helmholtz}. As noted already in \citep{dayan1995helmholtz},
a crucial drawback of the wake-sleep algorithm is that it postulates a pair forward (generation) and backward (recognition) models,
which together do not correspond to optimization of (a bound of) the marginal likelihood, because of the incorrect KL used therein, etc.
Thus the brain could not possibly be running such a algorithm, not even in principle! To the rescue, we note that the recent theory of
\textit{variational auto-encoders (VAEs)} \citep{kingma2013auto} might provide an efficient alternative to the wake-sleep algorithm,
as it overcomes the technical limits of the former, by using a reparametrization trick. For instance, unlike the wake-sleep algorithm
for minimizing free-energy, VAEs can be efficiently trained via back-propagation of prediction errors.

% \paragraph{Comparison to our proposed theory.}
On the surface, a common point between the FEP and our proposed RL-based framework
  is placing the minimization of a surprise signal at the core of brain function.
  Indeed in RL, surprise minimization is subsumed  by accurate prediction of
  rewarding outcomes in the future. A ``free-energy'' agent is barely a biomechanical machine which has the tendency to resist undesired phase-transitions
  \citep{friston2010free,fristonAIorRL,ortega2013thermodynamics}. Such a theory cannot by itself,
  explain the emergence of strategic behavior inherent in humans (e.g.,  cite dark-room experiment).

>>>>

\begin{equation}
  \pi^{\text{hard-max}}(s) = \argmax_{a \in \mathcal A}\tilde{Q}(s, a|\theta),
  \label{eq:policy}
\end{equation}
This defines a mapping from states directly to actions.
% $a \mapsto \tilde{Q}(a,s|\theta)$ of the $Q$-value function ``frozen'' at state $s$.
For instance, a simple linear model with a kernel $\phi$ would be of the
form $\tilde{Q}(s, a|\theta) = \phi(s,a)^T\theta$, where
$\phi(s,a)$ would represent a high-level representation of the state-action pairs
$(s,a)$, as was done proposed \citep{songNIPS2016}, or a eural network as recently proposed by in the
breakthrough works
\citep{mnih2015,silver2016mastering} for playing complex games (atari, Go, etc.) at superhuman levels.
%  and Parr et al. 2003 (JMLR)].
In the DMN, the dmPFC would implement such a hard-max lookup
over the action space.
The model
  parameters $\theta$ would correspond to synaptic weights and connection strengths between
  brain regions. It is a time-varying neuronal program which dictates how to move from world states $s$ to actions $a$ via the hard-max policy \eqref{eq:policy}.
  The approximating $Q$-value function $\tilde{Q}(s, a|\theta)$ would tell the DMN the (expected) usefulness of taking an action $a$ in state $s$.
  The DMN, and in particular its dmPFC node, could then contribute to the choice, at a given state $s$, of an action $a$ which maximizes these approximate
  $Q$-values. Such a mapping from states to actions is conventionally called \textit{policy matrix}\citep{mnih2015,silver2016mastering}.
  Learning consists in starting from a given table and
  updating it during action choices,
  which take the agent to different table entries.

@@

\subsection{The free-energy principle and active inference}
\textbf{XXX Passage: Explain FEP and active inference}
In Friston's free-energy principle (FEP) \citep{friston2010free} and active inference \citep{fristonAIorRL} theories, the brain is portrayed
as biomechanical inference engine
dedicated to minimizing the long-term average of surprise: the log-likelihood of the observe sensory input. It does this by actively generating hypothetical explanations or predictions of its sensory input.
Precursors of this theory can be traced back to \citep{dayan1995helmholtz} in which they introduced the so-called \textit{Helmholtz machine}, a hierarchical factorial \textit{directional deep belief-net}. According to the FEP account,
the goal of the brain is then to optimize over a generative model $G$ of sensations: to iteratively
modify its internal representation $p_G(\z|\x)$ about objects in the world, their interactions and dynamics,
so as to minimize surprise when these representations are confronted with sensory input $\x$ during perception cycles (i.e., the \textit{generative} model).
The FEP also postulates a dual model called the \textit{recognition} model,
which works in tandem with the generative model $p_R(\z|\x)$, to accomplish approximate inference. The recognition model dreams imaginary worlds $\z$,
while the latter tries to generate sensations $\x$ which match these dreams!

@@

The main algorithm for minimizing free-energy $F_G^R(\x)$ is the \textit{wake-sleep algorithm}
\citep{dayan1995helmholtz}. As noted already in \citep{dayan1995helmholtz},
a crucial drawback of the wake-sleep algorithm is that it postulates a pair forward (generation) and backward (recognition) models,
which together do not correspond to optimization of (a bound of) the marginal likelihood, because of the incorrect KL used therein, etc.
Thus the brain could not possibly be running such a algorithm, not even in principle! To the rescue, we note that the recent theory of
\textit{variational auto-encoders (VAEs)} \citep{kingma2013auto} might provide an efficient alternative to the wake-sleep algorithm,
as it overcomes the technical limits of the former, by using a reparametrization trick. For instance, unlike the wake-sleep algorithm
for minimizing free-energy, VAEs can be efficiently trained via back-propagation of prediction errors.

\subsubsection{RL reinterpretation of FEP and active inference}  FEP and active inference can be reinterpreted in our model-free RL framework by postulating that the reward signal maximized by the DMN is the negative of surprisal, i.e the log-likehood of sensory the current priors the agent has about the world. This is worth writing down explicitly,

\begin{equation}
  r^{\text{FEP}} = \underbrace{\log (p_G)}_{-\text{FEP generative surprisal}}.
    \end{equation}
% \paragraph{Comparison to our proposed theory.}
% On the surface, a common point between the FEP and our proposed RL-based framework
%   is placing the minimization of a surprise signal at the core of brain function.
%   Indeed in RL, surprise minimization is subsumed  by accurate prediction of
%   rewarding outcomes in the future. A ``free-energy'' agent is barely a biomechanical machine which has the tendency to resist undesired phase-transitions
%   \citep{friston2010free,fristonAIorRL,ortega2013thermodynamics}.
    Such a surprisal-based reinforcement learning scheme has been advocated for in \citep{schmidhuber2010,mohamed2015} under the equivalent framework of information compression.
    However, we note that though relevant, minimization of surprisal alone is insufficient to explain intelligent behavior observed in humans and other ``intelligent'' animals. The dark-room experiment, for example, provides a formidable argument against in favor of the insufficiency of the FEP and active inference theories to explain intelligent behavior in humans: when a human is put in a dark room, they don't just sit down in a corner forever, what would be the FEP-optimal thing to do; they actively search for the light switch, and start exploring the depths of the room, possibly looking for food or the way out!
